Sampling requirements to detect ecosystem change

### 2020-04-01 R/manuscript_draft.Rmd
edits to manuscript

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to true changepoint for total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling. Total number of samples was set between 5 and 30, out of the 74 initial time series.  \\label{fig:case_study_number_sample_change_biomass}'}
# Read data

@@

<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed. When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing. A decision must then be taken of which subsamples to analyze. To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig. \ref{fig:conceptual}). We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs. \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3). Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion.  (Fig. \ref{fig:simulation_min_time_plot}).

@@

Our approach goes beyond just paleoecological analyses. Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d,@White2020]. The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost. Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred. Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint. \easton{I'm not sure what this means. What do you mean by it ``matches"?}It matches very well the paleoecological approach [@Wingard2017]. However, both different types of data and different questions can be addressed with the same approach. Suppose instead that the goal was to detect a change in observed individuals with video-based approaches. It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest. Using an interval sampling approach, one could take a fixed number of samples to start. The trend over time from simple linear regression could be taken. Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates. If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample. Sampling would continue until the trend estimate reached an equilibrium. Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics. These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].

The approach is versatile: it also finds in 9 samples (11 to confirm) the change in species biomass (Fig. \ref{fig:case_study_number_sample_change_biomass}). The initial 5 samples analyzed are the same: a well tailored question from the start allows to reach the changepoint much more efficiently than random and regular subsampling methods. (\rosalie{do we want to include that? not sure})

>>>>

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to true changepoint for total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling. Total number of samples was set between 5 and 30, out of the 74 initial time series.  \\label{fig:case_study_number_sample}'}
# Read data

@@

<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed. When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing. A decision must then be taken of which subsamples to analyze. To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig. \ref{fig:conceptual}). We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs. \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3). Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion (Fig. \ref{fig:simulation_min_time_plot}).

@@

The approach is versatile: we converted the community data in a single vector of number of individuals per gram of sediment from the same sediment core (Fig. S5).  We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S6). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample} (c)). However, depending on the metric the user is interested in, the subsampling will target different subsamples to add. In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the begining.

Our approach goes beyond just paleoecological analyses. Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d,@White2020]. The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost. Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred. Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint. \easton{I'm not sure what this means. What do you mean by it ``matches"?} \rosalie{is it clearer that way? I changed "match" to "corresponds" and added more details.} It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occured, and allows subsampling at finer or rougher intervals [@Wingard2017]. However, both different types of data and different questions can be addressed with the same approach. Suppose instead that the goal was to detect a change in observed individuals with video-based approaches. It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest. Using an interval sampling approach, one could take a fixed number of samples to start. The trend over time from simple linear regression could be taken. Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates. If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample. Sampling would continue until the trend estimate reached an equilibrium. Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics. These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].
